---
title: "Introduction"
output: bookdown::html_document2
link-citations: yes
pkgdown:
  as_is: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lavaan.bingof)
```

```{r, echo = FALSE, results = "asis"}
# LaTeX shortcuts 
cat(readr::read_file("maths_shortcuts.tex"))
``` 

<!-- Extra LaTeX commands -->
\newcommand{\pimod}[1]{\pi_{#1}(\btheta)}
<!-- Extra LaTeX commands -->

Let $\bY = (Y_1, \dots, Y_p)^\top \in \{0,1\}^p$ be a vector of Bernoulli random variables. 
The probability of success for each Bernoulli component $Y_i$ is denoted $\dot\pi_i = \Pr(Y_i = 1)$, $i=1,\dots,p$.
Consider a response pattern $\by = (y_1,\dots,y_p)^\top$, where each $y_i\in\{0,1\}$.
The probability of observing such a response pattern is given by the joint distribution
\begin{align*}
\pi_{\by}
&= \Pr(\bY = \by)  = \Pr(Y_1=y_1,\dots,Y_p=y_p).
\end{align*}
Note that there are a total of $R=2^p$ possible joint probabilities corresponding to all possible two-way response patterns $\by$.
When we consider a parametric model with parameter vector $\btheta\in\bbR^q$, we write $\pi_{\by}(\theta)$ to indicate each joint probability, and 
\begin{equation}
\bpi(\theta) = \begin{pmatrix}
\pimod{1} \\
\vdots \\
\pimod{R}  \\
\end{pmatrix} \in [0,1]^R
(\#eq:jointprob)
\end{equation}
for the vector of joint probabilities, with $\sum_{r=1}^R \pimod{r} =1$.
