---
title: "Introduction"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
    number_sections: FALSE
link-citations: yes
bibliography: refs.bib
pkgdown:
  as_is: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lavaan.bingof)
```

```{r, echo = FALSE, results = "asis"}
# LaTeX shortcuts 
cat(readr::read_file("maths_shortcuts.tex"))
``` 

<!-- Extra LaTeX commands -->
\newcommand{\pimod}[1]{\pi_{#1}(\btheta)}
\newcommand{\Sigmaystar}{\bSigma_{\by^*}}
\newcommand{\pl}{\operatorname{\ell_P}}
\newcommand{\mlepl}{\hat\btheta_{\text{PL}}}
\newcommand{\mle}{\hat\btheta_{\text{ML}}}
\newcommand{\pimodpl}{\pi_{y_iy_j}^{(ij)}(\btheta)}
<!-- Extra LaTeX commands -->

Let $\bY = (Y_1, \dots, Y_p)^\top \in \{0,1\}^p$ be a vector of Bernoulli random variables. 
The probability of success for each Bernoulli component $Y_i$ is denoted $\dot\pi_i = \Pr(Y_i = 1)$, $i=1,\dots,p$.
Consider a response pattern $\by = (y_1,\dots,y_p)^\top$, where each $y_i\in\{0,1\}$.
The probability of observing such a response pattern is given by the joint distribution
\begin{align*}
\pi
&= \Pr(\bY = \by)  = \Pr(Y_1=y_1,\dots,Y_p=y_p).
\end{align*}
Note that there are a total of $R=2^p$ possible joint probabilities $\pi_r$ corresponding to all possible two-way response patterns $\by_r$.
When we consider a parametric model with parameter vector $\btheta\in\bbR^m$, we write $\pi_r(\btheta)$ to indicate each joint probability, and 
\begin{equation}
\bpi(\btheta) = \begin{pmatrix}
\pimod{1} \\
\vdots \\
\pimod{R}  \\
\end{pmatrix} \in [0,1]^R
(\#eq:jointprob)
\end{equation}
for the vector of joint probabilities, with $\sum_{r=1}^R \pimod{r} =1$.

## Model-based probabilities

The model of interest is a factor model, commonly used in social statistics.
Using an underlying variable (UV) approach, the observed binary responses $y_i$ are manifestations of some latent, continuous variables $Y_i^*$, $i=1,\dots,p$.
The connection is made as follows:
$$
Y_i = \begin{cases}
1 & Y_i^* > \tau_i \\
0 & Y_i^* \leq \tau_i,
\end{cases}
$$
where $\tau_i$ is the threshold associated with the variable $Y_i^*$.
For convenience, $Y_i^*$ is taken to be standard normal random variables^[For parameter identifiability, the location and scale of the normal distribution have to be fixed if the thresholds are to be estimated.].
The factor model takes the form
$$
\mathbf Y^* = \mathbf \Lambda\boldsymbol \eta + \boldsymbol \epsilon,
$$
where each component is explained below:

- $\mathbf Y^* = (Y_1^*,\dots,Y_p^*)^\top \in \mathbf R^p$ are the underlying variables;

- $\boldsymbol\Lambda \in \mathbf R^{p \times q}$ is the matrix of loadings;

- $\boldsymbol \eta = (\eta_1,\dots,\eta_q)^\top \in \mathbf R^q$ is the vector of latent factors;

- $\boldsymbol \epsilon \in \mathbf R^p$ are the error terms associated with the items (aka unique variables).

We also make some distributional assumptions, namely

1. $\boldsymbol\eta \sim \operatorname{N}_q(\mathbf 0, \boldsymbol\Psi)$, where $\bPsi$ is a correlation matrix, i.e. for $k,l\in\{1,\dots,q\}$,
$$
\bPsi_{kl} = \begin{cases}
1 & \text{if }  k = l \\
\rho(\eta_k, \eta_l) & \text{if } k \neq l.
\end{cases}
$$

2. $\boldsymbol \epsilon \sim \operatorname{N}_p(\mathbf 0, \boldsymbol\Theta_\epsilon)$, with $\boldsymbol\Theta_\epsilon = \mathbf I - \operatorname{diag}(\boldsymbol\Lambda \boldsymbol\Psi \boldsymbol\Lambda^\top)$.

These two assumptions, together with $\operatorname{Cov}(\boldsymbol\eta, \boldsymbol\epsilon) = 0$, implies that $\mathbf Y^*\sim\operatorname{N}_p(\mathbf 0,\Sigmaystar)$, where
\begin{align}
\Sigmaystar
= \operatorname{Var}(\mathbf Y^*) 
&= \boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^\top + \boldsymbol\Theta_\epsilon \\
&= \mathbf I + (\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^\top - \operatorname{diag}\big(\boldsymbol\Lambda \boldsymbol\Phi \boldsymbol\Lambda^\top)\big).
\end{align}
<!-- Evidently, -->
<!-- $$ -->
<!-- \operatorname{Cor}(y_i^*, y_{i'}^*) = \sum_{k,l=1}^q \lambda_{ik}\lambda_{i'l} (1 + \rho(z_k,z_l)). -->
<!-- $$ -->
The parameter vector for this factor model is denoted $\boldsymbol\theta^\top = (\boldsymbol\lambda, \boldsymbol\psi, \boldsymbol\tau) \in\bbR^m$, where it contains the vectors of the free non-redundant parameters in $\boldsymbol\Lambda$ and $\boldsymbol \Psi$ respectively, as well as the vector of all free thresholds.

Under this factor model, the probability of response pattern $\mathbf y_r$ is
\begin{align}
\pimod{r}
&= \Pr(\mathbf Y = \mathbf y_r \mid \boldsymbol\theta) \\
&= \idotsint_A \phi_p(\mathbf y^* \mid \mathbf 0, \boldsymbol\Sigma_{\mathbf y^*} ) \, \text{d}\mathbf y^*
\end{align}
where $\phi_p(\cdot \mid \boldsymbol\mu,\boldsymbol\Sigma)$ is the density function of the $p$-dimensional normal distribution with mean $\boldsymbol\mu$ and variance $\boldsymbol\Sigma$.
This integral is evaluated on the set
$$
A = \{ \mathbf Y^* \in \mathbb R^p \mid Y_1=y_1,\dots,Y_p=y_p \}.
$$


## Parameter estimation

Suppose that $h=1,\dots,n$ observations of $\bY=\by^{(h)}$ are obtained.
For the purpose of generalising from independent samples to complex samples, suppose that each unit $h$ in the sample is assigned a (normalised) survey weight $w_h$ with $\sum_h w_h = n$.
Of course, if an independent simple random sampling scheme is implemented, then each $w_h=1$.

The sample proportions for each category $r$ is written $p_r = \hat n_r/n$, where
<!-- Let $p_r = \hat n_r/n$ be the $r$th entry of the $R$-vector of proportions $\bp$, where -->
\begin{equation}
\hat n_r = \sum_h w_h [\by^{(h)} = \by_r],
\end{equation}
and $[\cdot]$ denotes the Iverson bracket.
In other words, $\hat n_r$ represents the (weighted) frequency counts of the observed response patterns such that $\sum_{r=1}^R \hat n_r = n$.
The vector $\hat\bn = (\hat n_1, \dots, \hat n_R)^\top$ then defines a *multivariate binomial distribution*, or more commonly called a multinomial distribution with parameters $n$, $R$, and $\bpi(\btheta)$.
The probability mass function of $\hat\bn$ is given by
\begin{equation}
f_{\hat\bn}(x_1,\dots,x_R) = N! \prod_{r=1}^R \frac{1}{x_r!} \big[\pimod{r}\big]^{x_r},
\end{equation}
and the log-likelihood for $\btheta$ given the observed frequencies $\hat\bn$ (ignoring constants) is then
\begin{equation}
\ell(\btheta) = \log f_{\hat\bn}(\hat n_1,\dots,\hat n_R) = \sum_{r=1}^{R} \hat n_r \log \pimod{r}.
\end{equation}

The maximum likelihood estimator $\mle$ satisfies $\mle = \argmax_{\btheta} \ell(\btheta)$.
Maximum likelihood theory tells us that, under certain regularity conditions, as $n\to\infty$,
\begin{equation}\label{eq:limitdisttheta}
\sqrt n (\hat\btheta - \btheta) \xrightarrow{\text D} \N\big(0, \cI_1(\btheta)\big),
\end{equation}
where $\cI_1(\btheta) \in \bbR^{m\times m}$ is the *unit Fisher information matrix*.
It can be further shown that $\cI_1(\btheta) = \bDelta^\top \bD^{-1} \bDelta$, where

- $\bDelta_{r,k} = \frac{\partial\pi_r(\btheta)}{\partial\theta_k}$, $r=1,\dots,R$, $k=1,\dots,t$; and
- $\bD = \diag(\bpi(\btheta))$.

Furthermore, the maximum likelihood estimators are a class of *best asymptotically normal* (BAN) estimators $\hat\btheta$ of $\btheta$ that satisfy
\begin{equation}
\sqrt n (\hat\btheta - \btheta) = \sqrt n \, \bB \big(\bp - \bpi(\btheta)\big) + o(1)
(\#eq:ban)
\end{equation}
for some $m\times R$ matrix $\bB$.
In the specific case of maximum likelihood estimators, we can derive $\bB$ to be $\bB = \cI^{-1} \bDelta^\top \bD^{-1}$.
This is proven in the corresponding article (see XXX for more details).

### Pairwise likelihood estimation

The main interest for this project is to construct test statistics using composite likelihood methods, specifically the pairwise likelihood method.
In order to define the pairwise likelihood, let $\pi_{y_iy_j}^{(ij)}(\btheta)$ be the probability under the model that $Y_i=y_i \in \{0,1\}$ and $Y_j=y_j\in\{0,1\}$ for a pair of variables $Y_i$ and $Y_j$, $i,j=1,\dots,p$ and $i<j$.
<!-- That is, -->
<!-- $$ -->
<!-- \pi_{y_iy_j}^{(ij)}(\btheta) = \Pr(Y_i = y_i, Y_j = y_j). -->
<!-- $$ -->
The pairwise log-likelihood takes the form
\begin{equation}
\pl(\btheta) = \sum_{i<j} \sum_{y_i}\sum_{y_j} \hat n_{y_iy_j}^{(ij)} \log \pi_{y_iy_j}^{(ij)}(\btheta),
\end{equation}
where $\hat n_{y_iy_j}^{(ij)}$ is the observed (weighted) frequency of sample units with $Y_i=y_i$ and $Y_j=y_j$,
$$
\hat n_{y_iy_j}^{(ij)} = \sum_h w_h [\by^{(h)}_i = y_i, \by^{(h)}_j = y_j].
$$
Let us also define the corresponding observed pairwise proportions $p_{y_iy_j}^{(ij)} = \hat n_{y_iy_j}^{(ij)}/n$.
There are a total of $\tilde R = 4 \times {p \choose 2}$ summands, where the '4' is representative of the total number of pairwise combinations of binary choices '00', '10', '01', and '11'.

The *pairwise maximum likelihood estimator* $\mlepl$ satisfies $\mlepl = \argmax_{\btheta} \pl(\btheta)$.
Under certain regularity conditions,
\begin{equation}
\sqrt N (\mlepl - \btheta) \xrightarrow{D} {\N}_t\big(\bzero, \cH(\btheta)\cJ(\btheta)^{-1}\cH(\btheta)\big),
\end{equation}
where
  
  - $\cH(\btheta)=-\E\nabla^2\pl(\btheta)$ is the *sensitivity matrix*; and
  - $\cJ(\btheta)=\Var\big(\sqrt N \nabla\pl(\btheta)\big)$ is the *variability matrix*.
  
We can also show that the pairwise maximum likelihood estimator satisfies \@ref(eq:ban), where the matrix $\bB$ in this case would be $\bB = \cH^{-1}\tilde\bDelta^{-1}\tilde \bD^{-1} \bG$, where 

- $\tilde\bDelta \in \bbR^{\tilde R \times t}$ consists of partial derivatives of the pairwise probabilities, i.e. $\frac{\partial\pimodpl}{\partial\theta_k}$;
- $\tilde\bD = \diag((\pimodpl)_{i<j})$; and 
- $\bG$ is some indicator matrix to transform the quantities from $\tilde R$ dimensions to $R$ dimensions. See other articles for further details.

## Goodness-of-fit test statistics

TBC




<!-- It is widely known [@agresti2012categorical] that -->
<!-- \begin{equation} -->
<!-- \sqrt N (\bp - \bpi) \xrightarrow{\text D} {\N}_R (\bzero, \bSigma), -->
<!-- \end{equation} -->
<!-- where $\bSigma = \bD - \bpi\bpi^\top$, with $\bD:= \diag(\bpi)$. -->





<!-- ## References -->













