---
title: "Estimating the multinomial covariance matrix"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
    number_sections: FALSE
link-citations: yes
bibliography: refs.bib
# nocite: |
#   @maydeu2008overview, @reiser1996analysis, @bartholomew2002goodness, @katsikatsou2012pairwise
pkgdown:
  as_is: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lavaan.bingof)
```

```{r, echo = FALSE, results = "asis"}
# LaTeX shortcuts 
cat(readr::read_file("maths_shortcuts.tex"))
``` 

<!-- Extra LaTeX commands -->
\newcommand{\pimod}[1]{\pi_{#1}(\btheta)}
\newcommand{\Sigmaystar}{\bSigma_{\by^*}}
\newcommand{\pl}{\operatorname{\ell_P}}
\newcommand{\mlepl}{\hat\btheta_{\text{PL}}}
\newcommand{\mle}{\hat\btheta_{\text{ML}}}
\newcommand{\pimodpl}{\pi_{y_iy_j}^{(ij)}(\btheta)}
\newcommand{\tr}{\operatorname{tr}}
<!-- Extra LaTeX commands -->

Let $\by$ be a $N \times S$ matrix, where $N$ is the total number of observations, and $S=p(p+1)/2$ is the total number of univariate plus bivariate responses.
That is, the columns represents observations from variables $y_1,\dots,y_p$ and $y_{12},y_{13},\dots,y_{p-1,p}$.
The matrix $\by$ contains 1's and 0's, indicating positive and negative responses respectively.

Each row vector $\by_i$, $i=1,\dots,N$ represents an observation from the multinomial distribution $\by_i \sim \Mult(\bp_2, \bSigma_2)$.
Under a complex sampling design, each observations has weight $w_i$.
For convenience, we normalise the sampling weights so that $\sum_{i=1}^N w_i = N$, so that each $w_i$ has the interpretation of an "effective sample size".

## Estimating the proportions

The weighted sample proportion estimator is 
$$
\hat\bp_2 = \frac{\sum_i w_i \by_i}{\sum_i w_i}
$$
which corresponds to a weighted column average of the matrix $\by$.

## Estimating the covariance matrix

The estimator for the covariance matrix is
$$
\hat\bSigma_2 = \frac{1}{\sum_i w_i} \sum_i \frac{N}{N-1} w_i (\by_i - \hat\bp_2)^\top (\by_i - \hat\bp_2).
$$
This means that each entry in this $S \times S$ covariance matrix is simply a weighted estimate of the variance or covariance of the binary response variables.
Under a model (e.g. binary factor model), we should replace the $\hat \bp_2$ with the estimated model probabilities $\hat\bpi_2 := \bpi_2(\hat\btheta)$ instead.

