---
output: pdf_document
toc: true
title: "Tables of simulation results"
pkgdown:
  as_is: true
  extension: pdf
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(tidyverse)
theme_set(theme_bw())
library(kableExtra)
library(lavaan.bingof)
load("simres_srs.RData")
load("simres_complex.RData")
```


## Simple random sampling

### Type I errors ($n=100$)

```{r}
srs_type1_res %>% 
  filter(n == 100) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage

### Type I errors ($n=250$)

```{r}
srs_type1_res %>% 
  filter(n == 250) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Type I errors ($n=500$)

```{r}
srs_type1_res %>% 
  filter(n == 500) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage

### Type I errors ($n=1000$)

```{r}
srs_type1_res %>% 
  filter(n == 1000) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Type I errors ($n=2000$)

```{r}
srs_type1_res %>% 
  filter(n == 2000) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Type I errors ($n=3000$)

```{r}
srs_type1_res %>% 
  filter(n == 3000) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```


### Power ($n=100$)

```{r}
srs_power_res %>% 
  filter(n == 100) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage

### Power ($n=250$)

```{r}
srs_power_res %>% 
  filter(n == 250) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Power ($n=500$)

```{r}
srs_power_res %>% 
  filter(n == 500) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage

### Power ($n=1000$)

```{r}
srs_power_res %>% 
  filter(n == 1000) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Power ($n=2000$)

```{r}
srs_power_res %>% 
  filter(n == 2000) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Power ($n=3000$)

```{r}
srs_power_res %>% 
  filter(n == 3000) %>%
  arrange(n, sim, name) %>%
  select(sim, name, everything(), -n, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage


## Complex sampling


### Type I errors (Stratified sampling)

```{r}
complex_type1_res %>% 
  filter(sampling == "Stratified") %>%
  arrange(sampling, sim, name) %>%
  select(sim, name, everything(), -sampling, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage

### Type I errors (Cluster sampling)

```{r}
complex_type1_res %>% 
  filter(sampling == "Cluster") %>%
  arrange(sampling, sim, name) %>%
  select(sim, name, everything(), -sampling, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Type I errors (Stratified cluster sampling)

```{r}
complex_type1_res %>% 
  filter(sampling == "Strat-clust") %>%
  arrange(sampling, sim, name) %>%
  select(sim, name, everything(), -sampling, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage


### Power (Stratified sampling)

```{r}
complex_power_res %>% 
  filter(sampling == "Stratified") %>%
  arrange(sampling, sim, name) %>%
  select(sim, name, everything(), -sampling, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```

\newpage

### Power (Cluster sampling)

```{r}
complex_power_res %>% 
  filter(sampling == "Cluster") %>%
  arrange(sampling, sim, name) %>%
  select(sim, name, everything(), -sampling, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```
\newpage

### Power (Stratified cluster sampling)

```{r}
complex_power_res %>% 
  filter(sampling == "Strat-clust") %>%
  arrange(sampling, sim, name) %>%
  select(sim, name, everything(), -sampling, -max_rank) %>%
  kbl(booktabs = TRUE, digits = 3) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  collapse_rows(1:2, row_group_label_position = "stack", latex_hline = "major")
  # collapse_rows(3:6, latex_hline = "major", valign = "top")
```


