---
title: "Description and results"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  out.width = "100%"
)
library(tidyverse)
theme_set(theme_bw())
library(kableExtra)
library(lavaan)
library(lavaan.bingof)
library(semPlot)
load("simplots.RData")
```

```{r, echo = FALSE, results = "asis"}
# LaTeX shortcuts 
cat(readr::read_file("maths_shortcuts.tex"))
``` 

A simulation study was carried out to investigate the performance of the proposed limited information goodness-of-fit tests.
The data was generated using an underlying variable framework.
The true parameter values were

- Loadings: $\boldsymbol\lambda = (0.8, 0.7, 0.47, 0.38, 0.34, \dots)$
- Factor correlations: $\phi = 0.3$ (1 factor) or $\bphi = (0.2, 0.3, 0.4)$ (2 factors)
- Thresholds: $\boldsymbol\tau = (-1.43, -0.55, -0.13, -0.82, -1.13,\dots)$

Five scenarios were investigated:

1. 1 factor, 5 variables
2. 1 factor, 8 variables
3. 1 factor, 15 variables
4. 2 factor, 10 variables
5. 3 factor, 15 variables

For each scenario, $B=1000$ data were generated either according to simple random sample or a complex sampling procedure using true parameter values, and the rejection rate (Type I error) were calculated.

To conduct a power analysis, an extra latent variable $x \sim \N(0,1)$ independent to the latent factor $\eta$ was added to the $y^*$ variables.
The loadings of $x$ are similar to the true values except that some noise ($\N(0,0.1^2)$) was added.
This means that the fitted model is misspecified because a missing factor was not accounted for.

The simulations took roughly 20 hours to complete.
No convergence issue reported when the sample size is in the range $500 \leq n \leq 3000$ (previous versions of this simulation study saw `{lavaan}` having difficulty converging when $n=100$ or $n=250$--these simulation results are not trustworthy so have been omitted).

## Path diagrams {.tabset}

### Model 1

```{r sem_path_mod1}
mod <- "
eta1  =~ y1 + y2 + y3 + y4 + y5
extra =~ y1  + y3 + y5
"
dat <- gen_data_bin(1, seed = 197)
fit <- sem(mod, dat, std.lv = TRUE) %>% suppressWarnings()

semPaths(fit, intercepts = FALSE, #residuals = TRUE, whatLabels = "est",  
         groups = list(c("eta1", paste0("y", c(2, 4))), c("extra")),
         manifests = paste0("y", 5:1),
         latents = c("extra", "eta1"),
         color = c("#28BBECFF", "#7A0403FF"),
         node.width = 0.9, nCharNodes = 0, sizeMan = 8, sizeLat = 9, 
         thresholds = FALSE, rotation = 3, exoCov = FALSE)
# "#30123BFF" "#28BBECFF" "#A2FC3CFF" "#FB8022FF" "#7A0403FF"
```

### Model 2

```{r sem_path_mod2}
mod <- "
eta1  =~ y1 + y2 + y3 + y4 + y5 + y6 + y7 + y8
extra =~ y1  + y3 + y5 + y7
# eta1 ~~ 0*extra
# eta1 ~~ 1*eta1
"
dat <- gen_data_bin(2, seed = 197)
fit <- sem(mod, dat, std.lv = TRUE) %>% suppressWarnings()

semPaths(fit, intercepts = FALSE, #residuals = TRUE, whatLabels = "est",  
         groups = list(c("eta1", paste0("y", c(2, 4, 6, 8))), c("extra")),
         manifests = paste0("y", 8:1),
         latents = c("extra", "eta1"),
         color = c("#28BBECFF", "#7A0403FF"),
         node.width = 0.9, nCharNodes = 0, sizeMan = 8, sizeLat = 9, 
         thresholds = FALSE, rotation = 3, exoCov = FALSE)
# "#30123BFF" "#28BBECFF" "#A2FC3CFF" "#FB8022FF" "#7A0403FF"
```

### Model 3

```{r sem_path_mod3}
mod <- "
eta1  =~ y1 +  y2 +  y3 +  y4 +  y5 + 
         y6 +  y7 +  y8 +  y9 + y10 +
        y11 + y12 + y13 + y14 + y15
extra =~ y1 + y3 +  y5 + 
         y7 + y9 + y11 +
        y13 + y15
# eta1 ~~ 0*extra
# eta1 ~~ 1*eta1
"
dat <- gen_data_bin(3, seed = 197)
fit <- sem(mod, dat, std.lv = TRUE) %>% suppressWarnings()

semPaths(fit, intercepts = FALSE, #residuals = TRUE, whatLabels = "est",  
         groups = list(c("eta1", paste0("y", c(2, 4, 6, 8))), c("extra")),
         manifests = paste0("y", 15:1),
         latents = c("extra", "eta1"),
         color = c("#28BBECFF", "#7A0403FF"),
         node.width = 0.9, nCharNodes = 0, sizeMan = 5, sizeLat = 9, 
         thresholds = FALSE, rotation = 3, exoCov = FALSE)
# "#30123BFF" "#28BBECFF" "#A2FC3CFF" "#FB8022FF" "#7A0403FF"
```

### Model 4

```{r sem_path_mod4}
mod <- "
eta1  =~ y1 +  y2 +  y3 +  y4 +  y5 
eta2  =~ y6 +  y7 +  y8 +  y9 + y10 
extra  =~ y1 +  y2 +  y3 +  y4 +  y5 + 
         y6 +  y7 +  y8 +  y9 + y10
eta1 ~~ 0*extra
eta2 ~~ 0*extra
# eta1 ~~ 0.3*eta2
"
dat <- gen_data_bin(4, seed = 197)
fit <- sem(mod, dat, std.lv = TRUE) %>% suppressWarnings()

semPaths(fit, intercepts = FALSE, what = "par", 
         whatLabels = "hide", #residuals = TRUE, whatLabels = "est",  
         groups = list(c("eta1", paste0("y", 1:5)), 
                       c("eta2", paste0("y", 6:10)), 
                       c("extra")),
         manifests = paste0("y", 10:1),
         latents = c("extra", "eta2", "eta1"),
         color = c("#30123BFF", "#A2FC3CFF", "#7A0403FF"),
         node.width = 0.9, nCharNodes = 0, sizeMan = 6, sizeLat = 9, 
         thresholds = FALSE, rotation = 3, weighted = FALSE)
```

### Model 5

```{r sem_path_mod5}
mod <- "
eta1  =~  y1 +  y2 +  y3 +  y4 +  y5 
eta2  =~  y6 +  y7 +  y8 +  y9 + y10 
eta3  =~ y11 + y12 + y13 + y14 + y15
extra  =~ y1 +  y2 +  y3 +  y4 +  y5 + 
          y6 +  y7 +  y8 +  y9 + y10 +
         y11 + y12 + y13 + y14 + y15
eta1 ~~ 0*extra
eta2 ~~ 0*extra
eta3 ~~ 0*extra
"
dat <- gen_data_bin(5, seed = 197)
fit <- sem(mod, dat, std.lv = TRUE) %>% suppressWarnings()

semPaths(fit, intercepts = FALSE, what = "par", 
         whatLabels = "hide", #residuals = TRUE, whatLabels = "est",  
         groups = list(c("eta1", paste0("y", 1:5)), 
                       c("eta2", paste0("y", 6:10)), 
                       c("eta3", paste0("y", 11:15)), 
                       c("extra")),
         manifests = paste0("y", 15:1),
         latents = c("extra", "eta3", "eta2", "eta1"),
         color = c("#30123BFF", "#1AE4B6FF", "#FABA39FF", "#7A0403FF"),
         node.width = 0.9, nCharNodes = 0, sizeMan = 5, sizeLat = 9, 
         thresholds = FALSE, rotation = 3, weighted = FALSE)
```

## SRS Type I errors {.tabset}

### 10% significance

```{r srs_type1a, fig.width = 7, fig.height = 5}
p_srs_a
```

### 5% significance {.active}

```{r srs_type1b, fig.width = 7, fig.height = 5}
p_srs_b
```

### 1% significance

```{r srs_type1c, fig.width = 7, fig.height = 5}
p_srs_c
```


## SRS Power plots  {.tabset}

### 10% significance

```{r srs_powerd, fig.width = 7, fig.height = 5}
p_srs_d
```

### 5% significance {.active}

```{r srs_powere, fig.width = 7, fig.height = 5}
p_srs_e
```

### 1% significance

```{r srs_powerf, fig.width = 7, fig.height = 5}
p_srs_f
```

## Complex sampling Type I errors {.tabset}

### 10% significance

```{r complex_type1a, fig.width = 7, fig.height = 8}
p_complex_a
```

### 5% significance {.active}

```{r complex_type1b, fig.width = 7, fig.height = 8}
p_complex_b
```

### 1% significance

```{r complex_type1c, fig.width = 7, fig.height = 8}
p_complex_c
```

## Complex sampling Power plots {.tabset}

### 10% significance

```{r complex_powerd, fig.width = 7, fig.height = 8}
p_complex_d
```

### 5% significance {.active}

```{r complex_powere, fig.width = 7, fig.height = 8}
p_complex_e
```

### 1% significance

```{r complex_powerf, fig.width = 7, fig.height = 8}
p_complex_f
```
