---
title: "Description and results"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  out.width = "100%"
)
library(tidyverse)
theme_set(theme_bw())
library(kableExtra)
library(lavaan.bingof)
load("simplots.RData")
```

```{r, echo = FALSE, results = "asis"}
# LaTeX shortcuts 
cat(readr::read_file("maths_shortcuts.tex"))
``` 

A simulation study was carried out to investigate the performance of the proposed limited information goodness-of-fit tests.
The data was generated using an underlying variable framework.
The true parameter values were

- Loadings: $\boldsymbol\lambda = (0.8, 0.7, 0.47, 0.38, 0.34, \dots)$
- Factor correlations: $\phi = 0.3$ (1 factor) or $\bphi = (0.2, 0.3, 0.4)$ (2 factors)
- Thresholds: $\boldsymbol\tau = (-1.43, -0.55, -0.13, -0.82, -1.13,\dots)$

Five scenarios were investigated:

1. 1 factor, 5 variables
2. 1 factor, 8 variables
3. 1 factor, 15 variables
4. 2 factor, 10 variables
5. 3 factor, 15 variables

For each scenario, $B=1000$ data were generated either according to simple random sample or a complex sampling procedure using true parameter values, and the rejection rate (Type I error) were calculated.

To conduct a power analysis, an extra latent variable $x \sim \N(0,1)$ independent to the latent factor $\eta$ was added to the $y^*$ variables.
The loadings of $x$ are similar to the true values except that some noise ($\N(0,0.1^2)$) was added.
This means that the fitted model is misspecified because a missing factor was not accounted for.

The simulations took roughly 20 hours to complete.
No convergence issue reported when the sample size is in the range $500 \leq n \leq 3000$ (previous versions of this simulation study saw `{lavaan}` having difficulty converging when $n=100$ or $n=250$--these simulation results are not trustworthy so have been omitted).

## SRS Type I errors {.tabset}

### 10% significance

```{r srs_type1a, fig.width = 7, fig.height = 5}
p_srs_a
```

### 5% significance {.active}

```{r srs_type1b, fig.width = 7, fig.height = 5}
p_srs_b
```

### 1% significance

```{r srs_type1c, fig.width = 7, fig.height = 5}
p_srs_c
```


## SRS Power plots  {.tabset}

### 10% significance

```{r srs_powerd, fig.width = 7, fig.height = 5}
p_srs_d
```

### 5% significance {.active}

```{r srs_powere, fig.width = 7, fig.height = 5}
p_srs_e
```

### 1% significance

```{r srs_powerf, fig.width = 7, fig.height = 5}
p_srs_f
```

## Complex sampling Type I errors {.tabset}

### 10% significance

```{r complex_type1a, fig.width = 7, fig.height = 8}
p_complex_a
```

### 5% significance {.active}

```{r complex_type1b, fig.width = 7, fig.height = 8}
p_complex_b
```

### 1% significance

```{r complex_type1c, fig.width = 7, fig.height = 8}
p_complex_c
```

## Complex sampling Power plots {.tabset}

### 10% significance

```{r complex_powerd, fig.width = 7, fig.height = 8}
p_complex_d
```

### 5% significance {.active}

```{r complex_powere, fig.width = 7, fig.height = 8}
p_complex_e
```

### 1% significance

```{r complex_powerf, fig.width = 7, fig.height = 8}
p_complex_f
```


