---
title: "Description and results"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(tidyverse)
theme_set(theme_bw())
library(kableExtra)
library(lavaan.bingof)
load("simres_srs.RData")
load("simres_complex.RData")
```

```{r, echo = FALSE, results = "asis"}
# LaTeX shortcuts 
cat(readr::read_file("maths_shortcuts.tex"))
``` 


A simulation study was carried out to investigate the performance of the proposed limited information goodness-of-fit tests.
The data was generated using an underlying variable framework.
The true parameter values were

- Loadings: $\boldsymbol\lambda = (0.8, 0.7, 0.47, 0.38, 0.34, \dots)$
- Factor correlations: $\phi = 0.3$ or $\bphi = (0.2, 0.3, 0.4)$
- Thresholds: $\boldsymbol\tau = (-1.43, -0.55, -0.13, -0.82, -1.13,\dots)$

Five scenarios were investigated:

1. 1 factor, 5 variables
2. 1 factor, 8 variables
3. 1 factor, 15 variables
4. 2 factor, 10 variables
5. 3 factor, 15 variables

For each scenario, $B=1000$ data were generated either according to simple random sample or a complex sampling procedure using true parameter values, and the rejection rate (Type I error) were calculated.

To conduct a power analysis, an extra latent variable $x \sim \N(0,1)$ independent to the latent factor $\eta$ was added to the $y^*$ variables.
The loadings of $x$ are similar to the true values except that some noise ($\N(0,0.1^2)$) was added.
This means that the fitted model is misspecified because a missing factor was not accounted for.

The simulations took roughly 20 hours to complete.

It is noted that model fit was problematic when sample size $n$ is small, and `{lavaan}` was either not able to converge or not able to invert the sensitivity matrix (or both). 
These simulations were then removed from the results.

# Plots

## SRS Type I errors

```{r srs_type1, fig.width = 8, fig.height = 5.5}
p_srs_a +   #theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))
p_srs_b +  # theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))
p_srs_c +   #theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))
```

## SRS Power plots

```{r srs_power, fig.width = 8, fig.height = 5.5}
p_srs_d +  # theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))
p_srs_e +  # theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))
p_srs_f + #  theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), shape = guide_legend(ncol = 2))
```

## Complex sampling Type I errors

```{r complex_type1, fig.width = 8, fig.height = 5.5}
p_complex_a #+   theme(legend.position = c(0.85, 0.15))
p_complex_b #+   theme(legend.position = c(0.85, 0.15))
p_complex_c #+   theme(legend.position = c(0.85, 0.15))
```

## Complex sampling Power plots

```{r complex_power, fig.width = 8, fig.height = 5.5}
p_complex_d +   #theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), alpha = guide_legend(ncol = 4))
p_complex_e +   #theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), alpha = guide_legend(ncol = 4))
p_complex_f +   #theme(legend.position = c(0.85, 0.15)) +
  guides(fill = guide_legend(ncol = 2), alpha = guide_legend(ncol = 4))
```
